Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records

### 2025-01-31 MANUSCRIPT.qmd
finalizing

@@

![Map of Arizona indicating US Geological Survey (USGS) streamgages used in this study. 8-digit HUC subbasin boundaries and physiographic regions shown.](images/StudyArea_20241203.png){#fig-study-area}

@@

While our data selection ensures a robust record for this analysis, it also highlights gaps in spatial coverage that the machine learning model can address. Arizona has 184 active USGS streamflow stations (as of 2024) across an area of 295,253 km², whereas Indiana (a humid state in the U.S.) has 189 active stations within an area of 94,326 km². This translates to a streamgage density of approximately 2.004 gages per 1,000 km² in Indiana, over three times higher than Arizona's density of 0.623 gages per 1,000 km². This stark difference underscores the relative sparsity of Arizona's streamgage network, particularly in the context of its larger geographic area and the unique hydrological challenges posed by its arid and semi-arid landscapes.

@@

Our initial training dataset comprised 7,724 observations across 45 variables. To optimize the model's performance, we first conducted an exhaustive grid search combined with 5-fold cross-validation to identify the optimal hyperparameter values. The hyperparameters evaluated included the learning rate ($\eta$), minimum split loss ($\gamma$), maximum tree depth, minimum child weight, and the number of trees. While not an exhaustive list of all possible XGBoost hyperparameters, the range of values explored provided sufficient variation to ensure the selection of a high-performing model.The optimal hyperparameters were determined to be: 700 trees, a learning rate ($\eta$) of 0.05, a minimum split loss ($\gamma$) of 0.075, a maximum tree depth of 7, and a minimum child weight of 5. Using these values, the XGBoost model was trained on the dataset with 10-fold cross-validation.

@@

The regionalized (HUC-8) long-term BFI (1991–2020) is shown in @fig-bfi-huc, while the long-term BFI for stream reaches with a Strahler stream order of 3 or greater is presented in @fig-bfi-streams. Basins with high BFIs, such as those along the Grand Canyon in the northwestern part of the study area, indicate greater surface water and groundwater interaction. Elevated BFI values are also observed along portions of the Mogollon Rim, a heavily forested region with high precipitation that marks the transition between physiographic regions. Additionally, headwater regions of perennial rivers tend to exhibit higher BFI values. In contrast, low BFI values are found in areas like the Defiance Plateau in northeastern Arizona and the arid southern regions of the state.

@@

![Predicted long-term BFI values for all stream reaches of Strahler stream order 3 and greater (1991-2020)](images/BFI_AllStream.png){#fig-bfi-streams}

@fig-bfi-streams indicates the predicted long-term BFI for stream reaches in Arizona of a Strahler stream order of 3 or greater. Stream-level BFI values often correspond to the long-term BFI observed in HUC-8 basins, as geospatial predictor data were aggregated at the basin level. @fig-bfi-streams highlights stream density across the region, revealing areas with higher concentrations of perennial streams, particularly in regions with favorable hydrogeologic conditions, such as spring-fed systems in the Verde River and the Salt River headwaters. Conversely, lower BFI values are observed in arid regions and those with lower stream densities such as parts of the Little Colorado River Basin, where groundwater recharge is limited [@flint_flint07]. While BFI values and their spatial patterns may align with certain terrain features or climatic patterns, it is important to recognize that BFI is driven by complex interactions among multiple factors [@fig-shap_values].

>>>>

  - name: Temuulen Sankey
    orcid: 0000-0002-6919-8428
    corresponding: false
    roles: 
      - Writing – Review & Editing 
    affiliations:
      - 'School of Informatics, Computing, and Cyber Systems, Northern Arizona University, Flagstaff, AZ, US'

@@

![Map of Arizona and US Geological Survey (USGS) streamgages used in this study. 8-digit HUC subbasin boundaries and physiographic regions shown.](images/StudyArea_20241203.png){#fig-study-area}

@@

While our data selection ensures a robust record for this analysis, it also highlights gaps in spatial coverage that the machine learning model can address. Arizona has 184 active USGS streamflow stations (as of 2024) across an area of 295,253 km², whereas Indiana (a humid state in the U.S.), for example, has 189 active stations within an area of 94,326 km². This translates to a streamgage density of approximately 2.004 gages per 1,000 km² in Indiana, over three times higher than Arizona's density of 0.623 gages per 1,000 km². This stark difference underscores the relative sparsity of Arizona's streamgage network, particularly in the context of its larger geographic area and the unique hydrological challenges posed by its arid and semi-arid landscapes.

@@

Our initial training dataset comprised 7,724 observations across 45 variables. To optimize the model's performance, we first conducted an exhaustive grid search combined with 5-fold cross-validation to identify the optimal hyperparameter values. The hyperparameters evaluated included the learning rate ($\eta$), minimum split loss ($\gamma$), maximum tree depth, minimum child weight, and the number of trees. While not an exhaustive list of all possible XGBoost hyperparameters, the range of values explored provided sufficient variation to ensure the selection of a high-performing model. The optimal hyperparameters were determined to be: 700 trees, a learning rate ($\eta$) of 0.05, a minimum split loss ($\gamma$) of 0.075, a maximum tree depth of 7, and a minimum child weight of 5. Using these values, the XGBoost model was trained on the dataset with 10-fold cross-validation.

@@

@@

The regionalized (HUC-8) long-term BFI (1991–2020) is shown in @fig-bfi-huc. Basins with high BFIs, such as those along the Grand Canyon in the northwestern part of the study area, indicate greater surface water and groundwater interaction. Elevated BFI values are also observed along portions of the Mogollon Rim, a heavily forested region with high precipitation that marks the transition between physiographic regions. Additionally, headwater regions of perennial rivers tend to exhibit higher BFI values. In contrast, low BFI values are found in areas like the Defiance Plateau in northeastern Arizona and the arid southern regions of the state. While BFI values and their spatial patterns may align with certain terrain features or climatic patterns, it is important to recognize that BFI is driven by complex interactions among multiple factors [@fig-shap_values].
